CKME 136 Final Project
Shabbir Yousuf Ali
Winter 2019

---
title: "Census Project"
output: pdf_document

---
1. Load requied libraries.
Install package install.packages("caret")
Install package install.packages("corrplot")
Install package install.packages('Boruta')


```{r library}
library(ggplot2)
library(corrplot)
library(Boruta)

library(randomForest)
library(ROCR)
library(caret)
library(rpart)
```

2. Load census data.
```{r census}
#setwd("c:/temp/cmk136/rawdata")
setwd("C:/Ryerson/ckme136/project/rawdata")
loc<-getwd()
censusdata <- read.csv(file="adult.data",header=TRUE,sep=",", na.string = "?")

#Add header to the columns
names(censusdata) <- c('age',
    'workclass',
    'fnlwgt',
    'education',
    'education-num',
    'marital-status',
    'occupation',
    'relationship',
    'race',
    'sex',
    'capital-gain',
    'capital-loss',
    'hours-per-week',
    'native-country',
    'income')
```

2.1. Divide the data into train and test data. 
```{r}
inTrain <- createDataPartition(y=censusdata$income, p= 0.75, list=FALSE)
training <- censusdata[inTrain,]
testing <- censusdata[-inTrain,]
```


3. Display dimensions, summary of data, names and overall structure of the data.

```{r }
data <- training
dim(data)
nrow(data)
ncol(data)
dim(testing)
summary(data)
names(data)
str(data)
```


4. Display Class Distributions.
```{r}
# Use the ggplot to find the income distribution <=50K VS >50K based on the training data
result = summary(data$income)/nrow(data) * 100
ggplot(data=data,aes(income)) + geom_bar(aes(fill = income), color = "black")
result 
```


5. Check and remove the missing values.

```{r}
cat("Missing values in training set:", sum(is.na(data)), "\n")
na_count <-sapply(data, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
na_count
nrow(data)
data <- na.omit(data)
nrow(data)
nrow(testing)
cat("Missing values in testing set:", sum(is.na(testing)), "\n")
na_count1 <-sapply(testing, function(y) sum(length(which(is.na(y)))))
na_count1
testingdata <- na.omit(testing)
nrow(testingdata)

```
5.1 Re-factoring the work class, occupation and native country after removing the NA values (exclude levels not required).

```{r}

data$workclass <- factor(data$workclass)
data$occupation <- factor(data$occupation)
data$native.country <- factor(data$`native-country`)
```

5.1 Re-factoring the work class, occupation and native country after removing the NA values (exclude levels not required) for testing data also.
```{r}

testingdata$workclass <- factor(testingdata$workclass)
testingdata$occupation <- factor(testingdata$occupation)
testingdata$native.country <- factor(testingdata$`native-country`)
```



6. Statistics of Numerical attributes

```{r}

#find the Min, Max, Mean, Median, 1st and 3rd Quarter of the numerical attributes
summary(data$age)
summary(data$`education-num`)
summary(data$`capital-gain`)
summary(data$`capital-loss`)
summary(data$`hours-per-week`)

# statistics of numerical attributes
summary(data$age)
sd(data$age)
hist(data$age, main = "Age Distribution",xlab = "Individual Age" ,col ="blue")
boxplot(data$age,main="Age ")
summary(data$education.num)
sd(data$education.num)
hist(data$`education-num`,main = "Education Distribution",xlab="Education in Years (yrs)",col = "blue")
boxplot(data$`education-num`,main="Education")
summary(data$`capital-gain`)
sd(data$`capital-gain`)
hist(data$`capital-gain`,main = "Capital Gain Distribution",xlab="Capital Gain",col = "blue")
boxplot(data$`capital-gain`,main="Capital Gain")
summary(data$`capital-loss`)
sd(data$`capital-loss`)
hist(log10(data$`capital-loss`),main = "Distribution of Capital Loss",xlab="Capital Loss",col = "blue")
boxplot(data$`capital-loss`,main="Capital Loss")
summary(data$`hours-per-week`)
sd(data$`hours.per.week`)
hist(data$`hours-per-week`,main = "Distribution of Hours Worked per Week",xlab="Hours worked per week",col = "blue")
boxplot(data$`hours-per-week`,main="Hours Worked per Week")

```

7. Fine the Correlation between numerical attributes.

```{r}
#Changing income to 0 <= 50k, 1 > 50k

data1 <- data
data1$income <- as.numeric(data1$income)-1
#Correlation plot
M <- c(1, 3, 5, 11:13, 15)
corrplot(cor(data1[,M]),method = "number")

##########################################################
# Correlations shows that numeric attributes are related #
# but are not strongly correlated. The variables are positively
# correlated. Education has the highest correlation 0.33 with income.
# Capital gain 0.22, age 0.24 and hours worked 0.23. The variables are positively correlted 
# with each other.
##########################################################
```

Saving the clean test and train data  in testdata.csv and traindata.csv respectively.
```{r}
traindata <- data
testdata  <- testingdata

write.csv(traindata, "traindata.csv", row.names = FALSE)
write.csv(testdata, "testdata.csv", row.names = FALSE)


```

```{r}
model <- glm(income ~ age+ workclass+ education+`marital-status`+ occupation+ sex +`hours-per-week`, data = traindata, family = binomial('logit'))
summary(model)

predicttrain <- predict(model,traindata,type='response')
pred1 <- rep('<=50K', length(predicttrain))
pred1[predicttrain>=.5] <- '>50K'
tb1 <- table(pred1, traindata$income)
tb1

```
